---
title: "Project Script"
output: html_document
---

Necessary Libraries (may need to install if not already done)
```{r}
library(ggmap)
library(ggplot2)
library(dplyr)
library(tidyverse)
library(sf)
library(mapview)
library(maps)
library(mapdata)
library(ggplot2)
library(sf)
library(viridis)
library(readxl)
```

Uploading and naming data. These are my pathways and will be different for each individual attempting recreation. DO NOT USE. The key thing here will be using the sheets = "" for your own pathway as the data set cannot be uploaded as a whole. Each sheet contains data from a different survey and cannot be distinguished if the entire data set is updated as one.  
```{r}
shrimpSM <- read_excel("Desktop/Data For Project.xlsx", 
    sheet = "Seamap")
View(Data_For_Project)

shrimpNM <- read_excel("Desktop/Data For Project.xlsx", 
    sheet = "Neamap")
View(Data_For_Project)

shrimpNM <- read_excel("Desktop/Data For Project.xlsx", 
    sheet = "Chesmap")
View(Data_For_Project)
```
Key points: anything with SM in the title is from Seamap, NM is from Neamap and CM is from Chesmmap. Keeping any names straight is essentia and keeping names the same with only the survey initials changing has made it easier to make code for one survey and easily edit to fit the others. 


Summarize Data
```{r}
summary(shrimpSM)
summary(shrimpNM)
summary(shrimpCM)
```

Map Survey Locations 
```{r}
##create a map of all the states to be included in the map
combinedmapregions <- map_data("state", region = c("florida","georgia",
        "south carolina","north carolina", "virginia","delaware", "maryland",
        "new jersey","pennsylvania","new york", "connecticut","rhode island", 
        "massachusetts", "ohio", "west virginia"))
```
```{r}
##create the map 
ggplot()+
geom_polygon(data = combinedmapregions, aes(x = long, y = lat, group = group), fill = "lightgray", color = "black") + ##creates an outline of the states
  geom_point(data = shrimpSM, 
                           aes(x = ELON, y = ELAT), fill="lightblue", color="blue", size=0.5, shape=21, stroke=0.25) +  ##includes seamap data on the map
   geom_point(data = shrimpNM, aes(x = lon, y = lat), fill="green", color="darkgreen", size=0.5, shape=21, stroke=0.25) + ##includes neamap data on the map
   geom_point(data = shrimpCM, aes(x = lon, y = lat), fill="yellow", color="orange", size=0.5, shape=21, stroke=0.25) + ##includes chesmmap data on the map 
  facet_wrap(vars(YEAR)) + ##creates a map for each year in the survey 
  theme_minimal() + 
  labs(title = "Shrimp Trawl Locations in the South Atlantic by Year", x = "Longitude (°W)", y = "Latitude (°N)") +
  theme(legend.position = "none",
        axis.text.x = element_text(size = 7, angle = 90, hjust = 1),
        axis.text.y = element_text(size = 7)) + ##sets theme specifics
  coord_cartesian(xlim = c(-81.5, -71.0), ylim = c(28.0, 41.5)) ##focuses the maps to be more centered around the data 
```

Map of where shrimp were actually caught and their abundances at those locations. 
```{r}
shrimpSMsf<- st_as_sf(shrimpSM, coords = c("ELON", "ELAT"), crs = 4326)
breaksSM <- seq(min(shrimpSM$TOWTOTAL, na.rm = TRUE), max(shrimpSM$TOWTOTAL, na.rm = TRUE), by = 20)
world_mapSM <- st_as_sf(maps::map("state", regions = c("florida","georgia","south carolina","north carolina"), 
                                  plot = FALSE, fill = TRUE))
shrimpSMsf$Abundance_Category <- cut(shrimpSMsf$TOWTOTAL, 
                                     breaks = c(0, 200, 400, 600, 800, 1000, 2000, 3000, 4000, Inf), 
                                     labels = c("0", "1-200", "201-400", "401-600", "601-800", "801-1000", "1001-2000", "2001-3000", "3001-4000+"))

shrimpSMsf_filtered <- shrimpSMsf %>% filter(Abundance_Category != 0)
shrimpSMabundance<-ggplot(shrimpSMsf_filtered) +
  geom_sf(data = world_mapSM, fill = "lightgray", color = "black") +
  geom_sf(aes(color = Abundance_Category), size = 0.5) + 
  facet_wrap(vars(YEAR))+
  scale_color_manual(values = c( "blue", "lightblue", "turquoise", "green", "darkgreen", "pink", "magenta", "red","purple"), 
                     name = "Abundance") +
  labs(title = "Shrimp Abundance Map SEAMAP", 
       x = "Longitude", 
       y = "Latitude", 
       color = "Abundance (TOWTOTAL)") +
  theme_minimal() + 
  theme(legend.position = "right",
        axis.text.x = element_text(size = 7, angle = 90, hjust = 1),
        axis.text.y = element_text(size = 9))+
  coord_sf(xlim = c(-81.5, -75.0), ylim = c(28.0, 36.0))
shrimpSMabundance

shrimpNMsf<- st_as_sf(shrimpNM, coords = c("lon", "lat"), crs = 4326)
breaksNM <- seq(min(shrimpNM$count, na.rm = TRUE), max(shrimpNM$count, na.rm = TRUE), by = 20)
world_mapNM <- st_as_sf(maps::map("state", regions = c("north carolina","virginia","delaware", "maryland","new jersey","pennsylvania","new york",                       "connecticut","rhode island", "massachusetts"), plot = FALSE, fill = TRUE))
shrimpNMsf$Abundance_Category <- cut(shrimpNMsf$count, 
                                     breaks = c(0, 200, 400, 600, 800, 1000, 2000, 3000, 4000, Inf), 
                                     labels = c("0", "1-200", "201-400", "401-600", "601-800", "801-1000", "1001-2000", "2001-3000", "3001-4000+"))

shrimpNMsf_filtered <- shrimpNMsf %>% filter(Abundance_Category != 0)
shrimpNMabundance <- ggplot(shrimpNMsf_filtered) +
  geom_sf(data = world_mapNM, fill = "lightgray", color = "black") +
  geom_sf(aes(color = Abundance_Category), size = 0.5) + 
  facet_wrap(vars(YEAR))+
  scale_color_manual(values = c("blue", "lightblue", "turquoise", "green", "darkgreen", "pink", "magenta", "red", "purple"), name = "Abundance") +
  labs(title = "Shrimp Abundance Map", 
       x = "Longitude", 
       y = "Latitude", 
       color = "Abundance (count)") +
  theme_minimal() + 
  theme(legend.position = "right",
        axis.text.x = element_text(size = 7, angle = 90, hjust = 1),
        axis.text.y = element_text(size = 9)) +
  coord_sf(xlim = c(-76.5, -70.0), ylim = c(35.0, 42.0))
shrimpNMabundance

shrimpCMsf<- st_as_sf(shrimpCM_clean, coords = c("lon", "lat"), crs = 4326)
breaksCM <- seq(min(shrimpCM$count, na.rm = TRUE), max(shrimpCM$count, na.rm = TRUE), by = 20)
world_mapCM <- st_as_sf(maps::map("state", regions = c("maryland","virginia","delaware"), plot = FALSE, fill = TRUE))
shrimpCMsf$Abundance_Category <- cut(shrimpCMsf$count, 
                                     breaks = c(0, 200, 400, 600, 800, 1000, 2000, 3000, 4000, Inf), 
                                     labels = c("0", "1-200", "201-400", "401-600", "601-800", "801-1000", "1001-2000", "2001-3000", "3001-4000+"))
shrimpCMsf_filtered <- shrimpCMsf %>% filter(Abundance_Category != 0)
shrimpCMabundance <- ggplot(shrimpCMsf_filtered) +
  geom_sf(data = world_mapCM, fill = "lightgray", color = "black") +
  geom_sf(aes(color = Abundance_Category), size = 1) + 
  facet_wrap(vars(year))+
  scale_color_manual(values = c("blue", "lightblue", "turquoise", "green", "darkgreen", 
                                "pink", "magenta", "red", "purple"), 
                     name = "Abundance") +
  labs(title = "L. setiferus Abundance Map (ChesMMAP)", 
       x = "Longitude", 
       y = "Latitude", 
       color = "Abundance (count)") +
  theme_minimal() + 
  theme(legend.position = "right",
        axis.text.x = element_text(size = 7, angle = 90, hjust = 1),
        axis.text.y = element_text(size = 7),
        panel.border = element_rect(color = "black", fill = NA, size = 1)) +
  coord_sf(xlim = c(-76.5, -75.0), ylim = c(36.5, 39.0))
shrimpCMabundance 
```

Calculate the total number caught per year 
```{r}
shrimpSMtotals <- aggregate(TOWTOTAL ~ YEAR, data = shrimpSM, sum)
shrimpSMtotals

shrimpNMtotals <- aggregate(count ~ YEAR, data = shrimpNM, sum)
shrimpNMtotals

shrimpCMtotals <- aggregate(count ~ YEAR, data = shrimpCM, sum)
shrimpCMtotals
```

Plot the totals by year 
```{r}
shrimpSMtotalplot<-ggplot(shrimpSMtotals, aes(x = YEAR, y = TOWTOTAL)) +
  geom_line(color = "purple", size = 1) +  
  labs(title = "SEAMAP L. setiferus",    
       x = "Year",                      
       y = "Total Number Caught") +                   
  theme_minimal()  
shrimpSMtotalplot

shrimpNMtotalsplot <-ggplot(shrimpNMtotals, aes(x = YEAR, y = count)) +
  geom_line(color = "purple", size = 1) +  
  labs(title = "NEAMAP L. setiferus",    
       x = "Year",                      
       y = "Total Number Caught") +                   
  theme_minimal()
shrimpNMtotalsplot

shrimpCMtotalsplot<- ggplot(shrimpCMtotals, aes(x = YEAR, y = count)) +
  geom_line(color = "purple", size = 1) +  
  labs(title = "CHESMMAP L. setiferus",    
       x = "Year",                      
       y = "Total Number Caught") +                   
  theme_minimal()  
shrimpCMtotalsplot
```
Percent Postive Catches
```{r}
#SEAMAP
positive_catch_by_yearSM <- shrimpSM %>%
  group_by(YEAR) %>%
  summarise(
    Total_Trawls = n(),
    Positive_Catches = sum(TOWTOTAL > 0, na.rm = TRUE),
    Percent_Positive = (Positive_Catches / Total_Trawls) * 100
  )
View(positive_catch_by_yearSM)
ggplot(positive_catch_by_yearSM, aes(x = YEAR, y = Positive_Catches)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "L. setiferus Catch by Year (SEAMAP)", x = "Year", y = "Positive Catches") +
  theme_minimal()+
  theme(
    panel.border = element_rect(colour = "black", fill = NA, size = 1),
  )
#NEAMAP
positive_catch_by_yearNM <- shrimpNM %>%
  group_by(YEAR) %>%
  summarise(
    Total_Trawls = n(),
    Positive_Catches = sum(count > 0, na.rm = TRUE),
    Percent_Positive = (Positive_Catches / Total_Trawls) * 100
  )
View(positive_catch_by_yearNM)
ggplot(positive_catch_by_yearNM, aes(x = YEAR, y = Positive_Catches)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "L. setiferus Catch by Year (NEAMAP)", x = "Year", y = "Positive Catches") +
  theme_minimal()+
  theme(
    panel.border = element_rect(colour = "black", fill = NA, size = 1),
  )

#ChesMMAP
positive_catch_by_yearCM <- shrimpCM %>%
  group_by(YEAR) %>%
  summarise(
    Total_Trawls = n(),
    Positive_Catches = sum(count > 0, na.rm = TRUE),
    Percent_Positive = (Positive_Catches / Total_Trawls) * 100
  )
View(positive_catch_by_yearCM)
ggplot(positive_catch_by_yearCM, aes(x = YEAR, y = Positive_Catches)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "L. setiferus Catch by Year", x = "Year", y = "Positive Catches") +
  theme_minimal()+
  theme(
    panel.border = element_rect(colour = "black", fill = NA, size = 1),
  )
```



Environmental Parameters: Temperature 
```{r}
##calculate averages per year
shrimpSMtemp<-aggregate(BTEMP ~ YEAR, data = shrimpSM, mean)
shrimpSMtemp

shrimpNMtemp<-aggregate(WT ~ YEAR, data = shrimpNM, mean)
shrimpNMtemp

shrimpCMtemp <-aggregate(WT~YEAR, data = shrimpCM, mean)
shrimpCMtemp
```
```{r}
##plot the averages per year 
shrimpSMtempplot <- ggplot(shrimpSMtemp, aes(x = YEAR, y = BTEMP)) +
  geom_line(color = "magenta", size = 1) +  
  labs(title = "SEAMAP Shrimp bottom temp",    
       x = "Year",                      
       y = "Average Bottom Tempurature") +                   
  theme_minimal() 
shrimpSMtempplot

shrimpNMtempplot <- ggplot(shrimpNMtemp, aes(x = YEAR, y = WT)) +
  geom_line(color = "magenta", size = 1) +  
  labs(title = "NEAMAP Shrimp bottom temp",    
       x = "Year",                      
       y = "Average Bottom Tempurature") +                   
  theme_minimal() 
shrimpNMtempplot

shrimpCMtempplot <- ggplot(shrimpCMtemp, aes(x = YEAR, y = WT)) +
  geom_line(color = "magenta", size = 1) +  
  labs(title = "ChesMMAP Shrimp bottom temp",    
       x = "Year",                      
       y = "Average Bottom Tempurature") +                   
  theme_minimal() 
shrimpCMtempplot
```

Environmental Parameters: Salinity 
```{r}
##calculate average salinity per year 
shrimpSMsalinity<-aggregate(BSAL ~ YEAR, data = shrimpSM, mean)
shrimpSMsalinity

shrimpNMsalinity<-aggregate(SA ~ YEAR, data = shrimpNM, mean)
shrimpNMsalinity

shrimpCMsalinity<-aggregate(SA~YEAR, data = shrimpCM, mean)
shrimpCMsalinity
```
```{r}
##plot average salinity per year 
shrimpSMsalinityplot <- ggplot(shrimpSMsalinity, aes(x = YEAR, y = BSAL)) +
  geom_line(color = "magenta", size = 1) +  
  labs(title = "SEAMAP Shrimp bottom salinity",    
       x = "Year",                      
       y = "Average Bottom Salinity") +                   
  theme_minimal() 
shrimpSMsalinityplot

shrimpNMsalinityplot <- ggplot(shrimpNMsalinity, aes(x = YEAR, y = SA)) +
  geom_line(color = "magenta", size = 1) +  
  labs(title = "NEAMAP Shrimp bottom salinity",    
       x = "Year",                      
       y = "Average Bottom Salinity") +                   
  theme_minimal() 
shrimpNMsalinityplot

shrimpCMsalinityplot <- ggplot(shrimpCMsalinity, aes(x = YEAR, y = SA)) +
  geom_line(color = "magenta", size = 1) +  
  labs(title = "ChesMMAP Shrimp bottom salinity",    
       x = "Year",                      
       y = "Average Bottom Salinity") +                   
  theme_minimal() 
shrimpCMsalinityplot
```
Statistical Analysis 
-assumptions typically tested for: linearity, normality, independence, constant variance, no multicollinerarity 
-in this case we know salinity and temperature are linked so we can assume collinearity between these two variables
-in larger data sets meeting these assumptions are less important
-the key ones here are going to be to test linearity and normality because this will tell us if we can use a type of linear regression or not as well as if the data may need to be transformed
-we are expecting to need to use a GAM based on similar studies but we'll try an LM first just to be sure
```{r}
##fit a model to get residuals 
##add the Towtotals from shrimpSMtotals to the shrimpSMtemp data table
shrimpSMtemp$TOWTOTAL <- shrimpSMtotals$TOWTOTAL
shrimpNMtemp$count <- shrimpNMtotals$count
shrimpCMtemp$count <- shrimpCMtotals$count
```

SEAMAP linear model for temperature
```{r}
##fit to a linear model and test residuals 
shrimpSMlmmodel <- lm(TOWTOTAL~BTEMP, data = shrimpSMtemp)
summary(shrimpSMlmmodel)
plot(shrimpSMlmmodel)
```

NEAMAP linear model for temperature
```{r}
shrimpNMlmmodel <- lm(count~WT, data = shrimpNMtemp)
summary(shrimpNMlmmodel)
plot(shrimpNMlmmodel)
```

ChesMMAP linear model for temperature
```{r}
shrimpCMlmmodel <- lm(count~WT, data = shrimpCMtemp)
summary(shrimpCMlmmodel)
plot(shrimpCMlmmodel)
```

Results of the above showed some non-normality (which was expected) so the data should be fit to a GAM model. 

SEAMAP generalized additive model for temperature
```{r}
library(mgcv)
shrimpSMtempgam <- gam(TOWTOTAL ~ s(BTEMP), data = shrimpSMtemp)
summary(shrimpSMtempgam)
```
NEAMAP generalized additive model for temperature
```{r}
shrimpNMtempgam <- gam(count ~ s(WT), data = shrimpNMtemp)
summary(shrimpNMtempgam)
```

ChesMMAP generalized additive model for temperature
```{r}
shrimpCMtempgam <- gam(count ~ s(WT), data = shrimpNMtemp)
summary(shrimpNMtempgam)
```
Repeat for salinity
```{r}
shrimpSMsalinity$TOWTOTAL <- shrimpSMtotals$TOWTOTAL
shrimpNMsalinity$count <- shrimpNMtotals$count
shrimpCMsalinity$count <- shrimpCMtotals$count
```

SEAMAP linear model for salinity
```{r}
shrimpSMlmsalinitymodel <- lm(TOWTOTAL~BSAL, data = shrimpSMsalinity)
summary(shrimpSMlmsalinitymodel)
plot(shrimpSMlmsalinitymodel)
```

NEAMAP linear model for salinity
```{r}
shrimpNMlmsalinitymodel <- lm(count~SA, data = shrimpNMsalinity)
summary(shrimpNMlmsalinitymodel)
plot(shrimpNMlmsalinitymodel)
```

ChesMMAP linear model for salinity
```{r}
shrimpCMlmsalinitymodel <- lm(count~SA, data = shrimpCMsalinity)
summary(shrimpCMlmsalinitymodel)
plot(shrimpCMlmsalinitymodel)
```

SEAMAP generalized additive model for salinity
```{r}
shrimpSMsalinitygam <- gam(TOWTOTAL ~ s(BSAL), data = shrimpSMsalinity)
summary(shrimpSMsalinitygam)
```

NEAMAP generalized additive model for salinity
```{r}
shrimpNMsalinitygam <- gam(count ~ s(SA), data = shrimpNMsalinity)
summary(shrimpNMsalinitygam)
```

ChesMMAP generalized additive model for salinity
```{r}
shrimpCMsalinitygam <- gam(count ~ s(SA), data = shrimpNMsalinity)
summary(shrimpNMsalinitygam)
```





















